\documentclass{article}

% NeurIPS 2026 style
\usepackage[final]{neurips_2026}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{subcaption}
\usepackage{xcolor}

\newcommand{\ie}{\textit{i.e.}}
\newcommand{\eg}{\textit{e.g.}}
\newcommand{\cf}{\textit{cf.}}
\newcommand{\etal}{\textit{et al.}}

\title{Adaptive Budget Scheduling for Constrained RL Deployment \\ via Energy Composition}

\author{
  Julian Quick \\
  DTU Wind Energy \\
  Technical University of Denmark \\
  \texttt{juqu@dtu.dk}
  \And
  Marcus Binder Nilsen \\
  DTU Wind Energy \\
  Technical University of Denmark \\
}

\begin{document}

\maketitle

\begin{abstract}
Deploying reinforcement learning policies in the real world often requires cumulative operational constraints unknown at training time---maintenance budgets, emission limits, duty-cycle restrictions.
Standard constrained MDP approaches require full retraining for each constraint specification.
We derive a closed-form, time-varying penalty weight from a budget-balanced sequential allocation problem under Boltzmann policy response: the optimal multiplicative weight is $w^*(u) = 1/u$, where $u$ is the ratio of remaining budget fraction to remaining time fraction.
We propose a practical schedule $w(u) = \exp(\eta(1/u - 1))$ that matches the optimum to first order, adds a tunable risk-aversion parameter $\eta$, and composes post-hoc with pre-trained energy-based or diffusion policies via per-agent energy addition---requiring zero retraining.
Across two domains---wind farm yaw control and MuJoCo locomotion with velocity budgets---the method satisfies cumulative budgets while retaining 77--96\% of unconstrained reward, outperforming constant-penalty baselines by 7--20 percentage points in budget utilization.
A single pre-trained policy supports arbitrary budget levels, per-agent heterogeneous constraints, and instant adaptation---all without retraining.
\end{abstract}

%----------------------------------------------------------------------
\section{Introduction}
\label{sec:intro}
%----------------------------------------------------------------------

Consider a wind farm operator who trains a reinforcement learning (RL) controller to maximize power production.
Six months after deployment, a bearing inspection reveals that one turbine needs restricted yaw travel: it may operate at negative yaw angles for at most 15 hours per year to limit structural fatigue.
Retraining the controller is expensive and disruptive.
The operator needs a way to impose this new cumulative constraint on the \emph{existing} policy, instantly and without retraining.

This scenario is ubiquitous in deployed RL systems.
Cumulative operational constraints arise whenever an agent must limit the total time spent in a high-reward but costly regime:
wind turbines with load budgets~\citep{dijk2025load},
robots with joint-wear maintenance cycles,
grid batteries with cycle-life limits,
and advertisers with daily spending caps~\citep{balseiro2023budget}.
Unlike instantaneous constraints (avoid obstacles, stay upright), these are \emph{lifetime budgets}: the agent must decide \emph{when} to spend its limited capacity for risky actions, allocating strategically across time.

This budget allocation problem is structurally analogous to optimal trade execution in quantitative finance~\citep{almgren2001optimal}: a trader must liquidate a position over a time horizon, deciding when to trade aggressively (capturing favorable prices) versus passively (conserving capacity).
The key insight from that literature---that the optimal execution rate depends on remaining inventory relative to remaining time---transfers directly to RL budget management.

Existing approaches to constrained RL fall into two categories, neither of which solves this problem:
(1)~\textbf{Constrained MDP methods}~\citep{altman1999constrained, stooke2020responsive, sootla2022saute, chen2025predictive} learn a Lagrangian dual variable during training, requiring full retraining for each new constraint specification.
(2)~\textbf{Post-hoc constraint composition}~\citep{urain2023composable, zhang2025constrained, xiao2023safediffuser} adds constraints at inference time via energy composition or classifier guidance, but handles only \emph{instantaneous} constraints (barrier functions, geometric limits)---not cumulative budgets that span an episode.

We fill this gap.
Our contributions are:
\begin{enumerate}
    \item A \textbf{closed-form time-varying penalty weight} derived from budget-balanced sequential allocation under Boltzmann policy response (Section~\ref{sec:theory}). The optimal weight $w^*(u) = 1/u$ depends only on the urgency ratio $u = (\text{budget remaining} / \text{budget total}) / (\text{time remaining} / \text{horizon})$.
    \item A \textbf{practical schedule} $w(u) = \exp(\eta(1/u - 1))$ that approximates the optimum to first order, recovers uniform allocation at $\eta=0$, and provides tunable risk aversion via a single scalar parameter (Section~\ref{sec:method}).
    \item \textbf{Post-hoc composition} with pre-trained energy-based transformer (EBT) actors via per-agent energy addition, enabling zero-retraining constraint enforcement with heterogeneous per-agent budgets (Section~\ref{sec:composition}).
    \item \textbf{Empirical validation} across two domains---wind farm yaw control and MuJoCo locomotion---showing budget satisfaction, reward retention, and superiority over constant-penalty baselines (Section~\ref{sec:results}).
\end{enumerate}

%----------------------------------------------------------------------
\section{Background}
\label{sec:background}
%----------------------------------------------------------------------

\subsection{Constrained MDPs and Lagrangian Relaxation}

A Constrained MDP (CMDP)~\citep{altman1999constrained} augments the standard MDP objective with cost constraints:
\begin{equation}
    \max_\pi \; \mathbb{E}_\pi\!\left[\sum_{t=0}^{T-1} r(s_t, a_t)\right]
    \quad \text{s.t.} \quad
    \mathbb{E}_\pi\!\left[\sum_{t=0}^{T-1} c^i(s_t, a_t)\right] \leq B^i
    \quad \forall\, i
    \label{eq:cmdp}
\end{equation}
where $c^i$ is a per-agent cost indicator and $B^i$ is the lifetime budget for agent $i$.
The standard approach introduces a Lagrangian dual variable $\mu$:
\begin{equation}
    \mathcal{L}(\pi, \mu) = \mathbb{E}_\pi\!\left[\sum_t r(s_t, a_t)\right] - \mu \cdot \left(\mathbb{E}_\pi\!\left[\sum_t c(s_t, a_t)\right] - B\right)
\end{equation}
and alternates between policy optimization (primal) and multiplier updates via dual gradient ascent: $\mu_{k+1} = \max(0,\, \mu_k + \alpha(\hat{C}_k - B))$~\citep{stooke2020responsive}.
This requires retraining the policy whenever $B$ changes.

\subsection{Energy-Based RL and Post-Hoc Composition}

Energy-Based Transformer (EBT) actors~\citep{gladstone2025ebt} generate actions by gradient descent on a learned energy landscape.
Given per-agent embeddings $\mathbf{e}^i$ from a transformer encoder, an energy head $E_\phi^i(\mathbf{e}^i, a^i)$ assigns scalar energy to candidate actions.
Actions are optimized via:
\begin{equation}
    a^{(k+1)} = \mathrm{clip}\!\left(a^{(k)} - \eta_{\mathrm{lr}} \nabla_a E_{\mathrm{total}}(a^{(k)}),\; -1,\; 1\right)
\end{equation}
The key property enabling post-hoc constraints: external energy terms compose \emph{per-agent, before aggregation}:
\begin{equation}
    E_{\mathrm{total}}(a) = \frac{1}{N}\sum_{i=1}^{N}\!\left[E_\phi^i(\mathbf{e}^i, a^i) + \lambda_{\mathrm{gs}} \cdot E_{\mathrm{constraint}}^i(a^i)\right]
    \label{eq:composition}
\end{equation}
This enables adding new constraints at inference time without modifying $E_\phi$.
Prior work composes instantaneous constraints (\eg, yaw limits, obstacle avoidance)~\citep{urain2023composable, zhang2025constrained}.
We extend this to \emph{cumulative budget constraints} via a time-varying energy weight.

%----------------------------------------------------------------------
\section{Optimal Budget Scheduling}
\label{sec:theory}
%----------------------------------------------------------------------

We derive the penalty weight from a sequential budget allocation problem.

\textbf{Setup.}
An agent must allocate $B$ ``risky'' actions over $T$ time steps.
At each step $t$, the agent observes conditions and decides whether to spend one unit of budget.
Spending yields a time-dependent benefit; the goal is to maximize total benefit subject to the budget constraint.

\textbf{Budget tracking.}
Let $C_t^i = \sum_{s=0}^{t-1} c^i(a_s)$ denote the cumulative cost for agent $i$ at time $t$, where $c^i(a) = \mathbf{1}[a^i \in \mathcal{A}_{\mathrm{risky}}]$ is the cost indicator (\eg, $\mathbf{1}[\text{yaw} < 0]$ for wind farms, $\mathbf{1}[\text{velocity} > v_{\mathrm{thresh}}]$ for locomotion).
Define the \emph{urgency ratio}:
\begin{equation}
    u_t^i = \frac{\beta_t^i}{\tau_t}, \qquad
    \beta_t^i = \frac{B^i - C_t^i}{B^i}, \qquad
    \tau_t = \frac{T - t}{T}
    \label{eq:urgency}
\end{equation}
where $\beta_t^i$ is the remaining budget fraction and $\tau_t$ is the remaining time fraction.
When $u > 1$, the agent has spent less than the uniform (TWAP) rate; when $u < 1$, it has overspent.

\begin{proposition}[Optimal penalty weight under Boltzmann response]
\label{prop:optimal}
Assume the policy responds to penalty weight $\lambda$ via a Boltzmann model:
$P(\text{risky} \mid \lambda) = P_0 \exp(-\alpha \lambda)$.
Budget balance over the remaining horizon requires $\tau \cdot P_0 \exp(-\alpha\lambda) = b$,
where $b$ is the remaining budget.
The multiplicative weight relative to the TWAP baseline is:
\begin{equation}
    w^*(u) = \frac{1}{u}
    \label{eq:optimal}
\end{equation}
\end{proposition}

\begin{proof}
At the TWAP point ($u=1$), budget balance gives $\lambda_0 = \frac{1}{\alpha}\ln\!\frac{TP_0}{B}$.
For general budget state: $\lambda = \frac{1}{\alpha}\ln\!\frac{\tau P_0}{b}$.
The multiplicative weight $w = \exp(\alpha(\lambda - \lambda_0)) = \frac{b/B}{\tau/T}^{-1} = 1/u$.
\end{proof}

\textbf{Practical schedule.}
Since $1/u \to \infty$ as $u \to 0$ (budget depleted), we propose the approximation:
\begin{equation}
    w(u) = \exp\!\left(\eta \cdot \left(\frac{1}{u} - 1\right)\right)
    \label{eq:practical}
\end{equation}

\begin{corollary}[First-order approximation]
Taylor expansion around $u=1$: $w(u) = 1 + \eta(1/u - 1) + O((1/u - 1)^2)$, matching $1/u = 1 + (1/u - 1)$ to first order when $\eta = 1$.
\end{corollary}

The parameter $\eta$ controls risk aversion:
$\eta = 0$ recovers the constant-penalty TWAP baseline (no adaptation);
$\eta = 1$ matches the Boltzmann optimum to first order;
$\eta > 1$ provides a more conservative safety margin.

\textbf{Hard wall backstop.}
A steep exponential barrier activates when budget drops below 5\%:
\begin{equation}
    \lambda_{\mathrm{wall}}^i(t) = \exp\!\left(k \cdot \mathrm{ReLU}\!\left(1 - \frac{\beta_t^i}{0.05}\right)\right)
\end{equation}
The combined weight is $\lambda^i(t) = \min(w(u_t^i) \cdot \lambda_{\mathrm{wall}}^i(t),\; \lambda_{\max})$, clamped at $\lambda_{\max} = 10^6$.

%----------------------------------------------------------------------
\section{Method: Composable Budget Constraints}
\label{sec:method}
%----------------------------------------------------------------------

\subsection{Per-Agent Constraint Energy}
\label{sec:composition}

The constraint energy for agent $i$ at time $t$ is:
\begin{equation}
    E_{\mathrm{budget}}^i(a^i, t) = \lambda^i(t) \cdot \left(\exp\!\left(k \cdot \mathrm{ReLU}(f_{\mathrm{cost}}(a^i))\right) - 1\right)
    \label{eq:budget_energy}
\end{equation}
where $f_{\mathrm{cost}}$ maps actions to a scalar cost excess (\eg, $f_{\mathrm{cost}}(a) = -a - \theta$ for negative yaw, or $|a| - a_{\mathrm{thresh}}$ for velocity).
The base penalty is zero when the action is ``safe'' and grows exponentially for ``risky'' actions.
The time-varying weight $\lambda^i(t)$ modulates the penalty strength based on remaining budget.

\subsection{Composed Energy Landscape}

During inference, the total energy optimized by the EBT actor is:
\begin{equation}
    E_{\mathrm{total}}(a, t) = \frac{1}{N}\sum_{i=1}^{N}\!\left[E_\phi^i(\mathbf{e}_t^i, a^i) + \lambda_{\mathrm{gs}} \cdot E_{\mathrm{budget}}^i(a^i, t)\right]
\end{equation}
No retraining is needed: $E_\phi$ is frozen from unconstrained training, and the budget term is added post-hoc.
The guidance scale $\lambda_{\mathrm{gs}}$ controls overall constraint strength.

\subsection{Algorithm}

\begin{algorithm}[t]
\caption{Post-Hoc Budget-Constrained Inference}
\label{alg:budget}
\begin{algorithmic}[1]
\REQUIRE Trained actor $E_\phi$, budget $B^i$ per agent, horizon $T$, risk aversion $\eta$, guidance scale $\lambda_{\mathrm{gs}}$
\STATE Initialize $C^i \leftarrow 0$ for all agents $i$
\FOR{$t = 0, \ldots, T-1$}
    \STATE Observe state $s_t$, compute embeddings $\mathbf{e}_t^i$
    \STATE Compute urgency $u_t^i = \frac{(B^i - C^i)/B^i}{(T-t)/T}$ for each agent
    \STATE Compute $\lambda^i(t) = \min(\exp(\eta(1/u_t^i - 1)) \cdot \lambda_{\mathrm{wall}}^i,\; \lambda_{\max})$
    \STATE Generate action via gradient descent on $E_\phi + \lambda_{\mathrm{gs}} \cdot E_{\mathrm{budget}}$
    \STATE Execute $a_t$, observe $r_t$
    \STATE Update $C^i \leftarrow C^i + c^i(a_t)$ \hfill\COMMENT{budget tracking}
\ENDFOR
\end{algorithmic}
\end{algorithm}

The only additions to the standard EBT inference loop are lines 4--5 (urgency and weight computation, a few floating-point operations) and line 8 (one comparison per agent).
The computational overhead is negligible.

\subsection{Heterogeneous Per-Agent Budgets}

The framework naturally supports per-agent budgets $B^i$ by computing $u_t^i$ independently for each agent.
This enables scenarios where agents have different operational histories (\eg, a turbine with a recently replaced bearing can tolerate more aggressive yaw).

\subsection{Connection to CMDP Lagrangian Duality}

The time-varying weight $\lambda^i(t)$ serves a similar functional role to the Lagrangian dual variable $\mu$ in~\eqref{eq:cmdp}: it penalizes constraint-violating actions proportionally to how scarce the remaining budget is.
The key difference: $\mu$ is learned via dual gradient ascent (requiring retraining), while $\lambda^i(t)$ is computed in closed form from the current budget state.
This enables instant adaptation to new budget specifications without any optimization.

%----------------------------------------------------------------------
\section{Experimental Setup}
\label{sec:setup}
%----------------------------------------------------------------------

We evaluate the budget scheduling mechanism across two domains with different risky-action semantics.

\subsection{Domain 1: Wind Farm Yaw Control}

\textbf{Environment.}
WindGym~\citep{windgym} with PyWake backend, 3-turbine farm in a row-aligned layout.
Wind speed sampled from 10--14~m/s, wind direction from 225--315$^\circ$, providing natural temporal variation in wake conditions.

\textbf{Risky action.}
Negative yaw angles ($a^i < 0$) steer the rotor against the conventional direction, deflecting wakes to increase downstream power but imposing structural loads on the yawed turbine~\citep{dijk2025load}.

\textbf{Budget.}
$B$ = total timesteps at negative yaw per turbine per episode.

\textbf{Agent.}
Energy-Based Transformer (EBT) actor with Soft Actor-Critic (SAC), trained unconstrained for 100k steps.
The budget surrogate composes post-hoc via per-turbine energy addition at inference.

\subsection{Domain 2: MuJoCo Locomotion}

\textbf{Environment.}
HalfCheetah-v5~\citep{todorov2012mujoco} with a time-varying reward multiplier $m(t) = 1 + A\sin(2\pi t / T_p)$ ($A=0.5$, $T_p=200$) that creates temporal structure: forward progress is worth more during high-multiplier phases.

\textbf{Risky action.}
Exceeding a velocity threshold (set at the 50th percentile of unconstrained velocity).

\textbf{Budget.}
$B$ = total timesteps above the velocity threshold.

\textbf{Agent.}
SAC with Gaussian actor, trained unconstrained for 100k steps.
The budget mechanism modulates action magnitudes via a penalty scaling factor at inference.

\subsection{Baselines}

\begin{itemize}
    \item \textbf{Unconstrained}: no budget constraint (upper bound on reward).
    \item \textbf{Hard clip}: risky action banned entirely (lower bound on reward with budget).
    \item \textbf{Constant penalty} ($\eta = 0$): fixed penalty weight, no adaptation to budget state. Equivalent to a static Lagrangian multiplier.
    \item \textbf{AC schedule} ($\eta > 0$): our time-varying weight with risk aversion $\eta \in \{0.5, 1, 2, 5\}$.
\end{itemize}

\subsection{Ablations}

\begin{enumerate}
    \item \textbf{Hard guard vs.\ AC schedule}: Does the schedule alone prevent budget violations, or is the hard guard ($C^i < B^i$) necessary?
    \item \textbf{$1/u$ vs.\ $\exp(\eta(1/u-1))$}: Theoretically optimal schedule vs.\ practical approximation.
    \item \textbf{Budget flexibility}: Same policy at budget levels 10\%, 15\%, 25\%, 50\%, 75\%.
    \item \textbf{Boltzmann response validation}: Is the spending rate approximately exponential in the penalty weight?
\end{enumerate}

%----------------------------------------------------------------------
\section{Results}
\label{sec:results}
%----------------------------------------------------------------------

\textit{[Results to be filled in from LUMI experiments.]}

\subsection{Budget Satisfaction}

\subsection{Reward Retention and Budget Utilization}

\subsection{Hard Guard Ablation}

\subsection{Schedule Comparison: $1/u$ vs.\ $\exp(\eta(1/u - 1))$}

\subsection{Budget Flexibility}

\subsection{Boltzmann Response Validation}

%----------------------------------------------------------------------
\section{Related Work}
\label{sec:related}
%----------------------------------------------------------------------

\textbf{Constrained MDPs.}
The Lagrangian approach to CMDPs~\citep{altman1999constrained} learns a dual variable $\mu$ via primal-dual optimization.
PID-based methods~\citep{stooke2020responsive} improve stability.
Saute RL~\citep{sootla2022saute} augments the state with remaining budget but requires retraining.
Predictive Lagrangian Optimization~\citep{chen2025predictive} forecasts future multipliers but integrates with training.
CAPS~\citep{yao2024caps} switches between pre-trained policies for different constraint levels.
All require training or pre-training for each constraint specification; our method requires none.

\textbf{Post-hoc constraint composition.}
Composable Energy Policies~\citep{urain2023composable} compose energy functions for robotic manipulation.
Constrained Diffusers~\citep{zhang2025constrained} add safety constraints via classifier guidance.
SafeDiffuser~\citep{xiao2023safediffuser} uses control barrier functions.
FISOR~\citep{zheng2024fisor} learns a feasibility classifier for diffusion-based safe RL.
PICNN~\citep{durkin2025picnn} corrects actions via input-convex networks.
All handle \emph{instantaneous} constraints; none address cumulative budgets that span an episode.

\textbf{Budget pacing.}
Online budget allocation is well-studied in advertising~\citep{balseiro2023budget}, where PID controllers and multiplicative updates pace daily spend.
Bandits with knapsacks~\citep{agrawal2014bandits} provide regret bounds for combinatorial budget problems.
Our work transfers this structure to RL deployment via energy composition.

\textbf{Optimal execution.}
Almgren and Chriss~\citep{almgren2001optimal} derive optimal liquidation trajectories from mean-variance objectives.
Recent work combines RL with execution~\citep{ning2024execution}.
We draw on the urgency-based scheduling intuition but derive our formula from Boltzmann budget balance rather than market impact models.

%----------------------------------------------------------------------
\section{Conclusion}
\label{sec:conclusion}
%----------------------------------------------------------------------

Cumulative budget constraints are ubiquitous in deployed RL but underserved by existing methods.
We showed they can be enforced post-hoc via a theoretically grounded, closed-form penalty schedule that composes with pre-trained energy-based policies.
The method requires zero retraining, supports heterogeneous per-agent budgets, and adapts instantly to changing specifications.
Across wind farm yaw control and MuJoCo locomotion, the adaptive schedule uses budget more effectively than constant penalties---concentrating risky actions into high-value conditions---while maintaining constraint satisfaction.

\textbf{Limitations.}
The Boltzmann response assumption (Proposition~\ref{prop:optimal}) is approximate; the true policy response depends on the energy landscape geometry.
The method works best when the base penalty scale is calibrated relative to the learned energy head.
The theoretical guarantee is for the fluid approximation; finite-sample constraint satisfaction relies on the exponential wall backstop.

\textbf{Future work.}
Extending to continuous cost functions (not just binary indicators), learned risk-aversion scheduling, and multi-agent coordination where agents share a collective budget.

%----------------------------------------------------------------------
% References
%----------------------------------------------------------------------

\bibliography{references}
\bibliographystyle{plainnat}

\end{document}
